Collective SU($n$) spin model dynamics

### 2020-07-30 sun_phases/sun_dynamics_notes.tex
changed conventions for order parameter and add a small-angle approximation

@@

Consider the single-particle Hamiltonian
\begin{align}
  H = -\f{J}{2} \sum_{\mu,q} \cos\p{q + \mu\phi} s_{\mu\mu q},
\end{align}
where $q$ is a quasi-momentum (in units with lattice spacing $a=1$);
$\mu\in\set{-I,-I+1,\cdots,I}$ is a nuclear spin projection onto a
quantization axis, with $I\equiv\p{n-1}/2$ the total nuclear spin;
$\phi$ is a spin-orbit-coupling angle, and $J$ is the single-particle
bandwidth (proportional to the tunneling rate, $J/4$).  We wish to
compute an expectation value of the operator
\begin{align}
  S^2\p{t}
  \equiv e^{it H} \p{\v S^\dag \c\v S} e^{-it H}
  = \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q} \times
  e^{itJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}},
\end{align}
where
\begin{align}
  \eta_{\mu\nu p}
  \equiv -\f12 \sp{\cos\p{p+\mu\phi} - \cos\p{p+\nu\phi}}
  = \sin\p{\phi_{\mu\nu}^-} \sin\p{p+\phi_{\mu\nu}^+},
  &&
  \phi_{\mu\nu}^\pm \equiv \f{\mu\pm\nu}{2}\,\phi.
\end{align}
If $\ket\Psi=\ket\psi^{\otimes N}$, then
\begin{align}
  \bk{\Psi|S^2\p{t}|\Psi}
  &= \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \sum_{p\ne q} e^{itJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}} + n N \\
  &= \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \sum_{p,q} e^{itJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}} + \p{n-1} N \\
  &= N^2 \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \abs{\xi_{\mu\nu}\p{\phi}}^2 + \p{n-1} N,
\end{align}
where
\begin{align}
  \xi_{\mu\nu}\p{\phi} \equiv \f1N \sum_q e^{itJ\eta_{\mu\nu q}}
  = \f1N \sum_q e^{itJ\sin\p{\phi_{\mu\nu}^-} \sin\p{q+\phi_{\mu\nu}^+}}.
\end{align}
As $N\to\infty$, this sum is well approximated by an integral:
\begin{align}
  \xi_{\mu\nu}\p{\phi} \approx \f1{2\pi} \int_{-\pi}^\pi \dd\theta\,
  e^{itJ\sin\p{\phi_{\mu\nu}^-} \sin\p{\theta+\phi_{\mu\nu}^+}}
  = \J_0\sp{tJ\sin\p{\phi_{\mu\nu}^-}},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.  To
leading order in $N\gg1$, we thus find that
\begin{align}
  \bk{\Psi|S^2\p{t}|\Psi}
  \approx N^2 \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \J_0\sp{tJ\sin\p{\phi_{\mu\nu}^-}}^2.
\end{align}
The time-independent terms in the sum have $\mu=\nu$ (in which case
$\phi_{\mu\nu}^-=0$), so
\begin{align}
  \bk{\Psi|S^2_{\t{const}}|\Psi}
  \approx N^2 \sum_{\mu,\mu} \abs{\bk{\psi\op{\mu}{\mu}\psi}}^2
  = N^2 r,
  &&
  r \equiv \sum_\mu \abs{\bk{\mu|\psi}}^4.
  \label{eq:S^2_const}
\end{align}
Up to an inconsequential choice of gauge, we can prepare a state of
the form
\begin{align}
  \ket\psi = \f1{\sqrt{2}}
  \p{e^{i\beta} \ket{\theta,\alpha}
    + e^{-i\beta} \ket{\theta,-\alpha}},
  &&
  \ket{\theta,\alpha} \equiv e^{i\alpha s_\z} e^{i\theta s_\y} \ket\up,
\end{align}
where $\theta,\alpha$ are respectively polar and azimuthal angles of
the polarized state $\ket{\theta,\alpha}$ of a single spin; the spin
operators $s_\z,s_\y$ generate rotations about the $z,y$ axes, with
e.g.~$s_\z\equiv\sum_\mu \mu\op\mu$; and $\ket\up$ is a state
polarized along the $z$ axis.  More specifically, the state that we
can prepare has
\begin{align}
  \theta = \f{\pi}{2} + \arcsin\p{\f13},
  &&
  \alpha = \f{\pi}{3},
  &&
  \beta = -\f{2\pi}{3}\, I.
\end{align}
In order to evaluate \eqref{eq:S^2_const}, we therefore
expand\footnote{The expansion in \eqref{eq:polarized_proj} is acquired
  by representing an $n$-level spin by $n-1$ qubits restricted to the
  (Dicke) manifold of permutationally symmetric states.}
\begin{align}
  \bk{\mu|\theta,\alpha} = { 2I \choose I+\mu }^{1/2}
  \cos\p{\f{\theta}{2}}^{I+\mu} \sin\p{\f{\theta}{2}}^{I-\mu}
  e^{-i\mu\alpha},
  \label{eq:polarized_proj}
\end{align}
which implies that
\begin{align}
  r = \sum_\mu { 2I \choose I+\mu }^2
  \sp{\cos\p{\f{\theta}{2}}^{I+\mu} \sin\p{\f{\theta}{2}}^{I-\mu}}^4
  \times 4 \cos\p{\mu\alpha-\beta}^4.
\end{align}
Figure \ref{fig:const_vals} shows several values of $r$ for different
$n$.

\begin{figure}
  \centering
  \includegraphics{figures/oscillations/const_vals.pdf}
  \caption{Several values of
    $r \equiv\sum_\mu\abs{\bk{\mu|\psi}}^4 =
    \lim_{N\to\infty}\bk{\Psi|S^2_{\t{const}}|\Psi}/N^2$ for different
    $n$.}
  \label{fig:const_vals}
\end{figure}

>>>>

\let\var\relax
\DeclareMathOperator{\var}{var}

@@

Consider the single-particle Hamiltonian
\begin{align}
  H = -\f{J}{2} \sum_{\mu,q} \cos\p{q + \mu\phi} s_{\mu\mu q},
\end{align}
where $q$ is a quasi-momentum (in units with lattice spacing $a=1$);
$\mu\in\set{-I,-I+1,\cdots,I}$ is a nuclear spin projection onto a
quantization axis, with $I\equiv\p{n-1}/2$ the total nuclear spin;
$\phi$ is a spin-orbit-coupling angle, and $J$ is the single-particle
bandwidth (proportional to the tunneling rate, $J/4$).  We wish to
compute an expectation value of the operator
\begin{align}
  \O\p{t}
  \equiv e^{it H} \p{\f{\v S^\dag \c\v S}{N^2}} e^{-it H}
  = \f1{N^2} \sum_{\mu,\nu,p,q} s_{\mu\nu p} s_{\nu\mu q} \times
  e^{itJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}},
\end{align}
where
\begin{align}
  \eta_{\mu\nu p}
  \equiv -\f12 \sp{\cos\p{p+\mu\phi} - \cos\p{p+\nu\phi}}
  = \sin\p{\phi_{\mu\nu}^-} \sin\p{p+\phi_{\mu\nu}^+},
  &&
  \phi_{\mu\nu}^\pm \equiv \f{\mu\pm\nu}{2}\,\phi.
\end{align}
If $\ket\Psi=\ket\psi^{\otimes N}$, then
\begin{align}
  \bk{\Psi|\O\p{t}|\Psi}
  &= \f1{N^2} \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \sum_{p\ne q} e^{itJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}} + \f{n}{N} \\
  &= \f1{N^2} \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \sum_{p,q} e^{itJ\p{\eta_{\mu\nu p}-\eta_{\mu\nu q}}} + \f{n-1}{N} \\
  &= \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \abs{\xi_{\mu\nu}\p{\phi}}^2 + \f{n-1}{N},
\end{align}
where
\begin{align}
  \xi_{\mu\nu}\p{\phi} \equiv \f1N \sum_q e^{itJ\eta_{\mu\nu q}}
  = \f1N \sum_q e^{itJ\sin\p{\phi_{\mu\nu}^-} \sin\p{q+\phi_{\mu\nu}^+}}.
\end{align}
As $N\to\infty$, this sum is well approximated by an integral:
\begin{align}
  \xi_{\mu\nu}\p{\phi} \approx \f1{2\pi} \int_{-\pi}^\pi \dd\theta\,
  e^{itJ\sin\p{\phi_{\mu\nu}^-} \sin\p{\theta+\phi_{\mu\nu}^+}}
  = \J_0\sp{tJ\sin\p{\phi_{\mu\nu}^-}},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.  To
leading order in $N\gg1$, we thus find that
\begin{align}
  \bk{\Psi|\O\p{t}|\Psi}
  \approx \sum_{\mu,\nu} \abs{\bk{\psi\op{\mu}{\nu}\psi}}^2
  \J_0\sp{tJ\sin\p{\phi_{\mu\nu}^-}}^2.
  \label{eq:O_time}
\end{align}

\subsection{Long-time average}

The time-independent terms in the sum have $\mu=\nu$ (in which case
$\phi_{\mu\nu}^-=0$), so
\begin{align}
  \bk{\Psi|\O_{\t{const}}|\Psi}
  \approx r \equiv \sum_\mu \abs{\bk{\mu|\psi}}^4.
  \label{eq:O_const}
\end{align}
where $r$ is sometimes called the inverse participation ratio of
$\ket\psi$ with respect to the basis $\set{\ket\mu}$.  Up to an
inconsequential choice of gauge, we can prepare a state of the form
\begin{align}
  \ket\psi = \f1{\sqrt{2}}
  \p{e^{i\beta} \ket{\theta,\alpha}
    + e^{-i\beta} \ket{\theta,-\alpha}},
  &&
  \ket{\theta,\alpha} \equiv e^{i\alpha s_\z} e^{i\theta s_\y} \ket\up,
\end{align}
where $\theta,\alpha$ are respectively polar and azimuthal angles of
the polarized state $\ket{\theta,\alpha}$ of a single spin; the spin
operators $s_\z,s_\y$ generate rotations about the $z,y$ axes, with
e.g.~$s_\z\equiv\sum_\mu \mu\op\mu$; and $\ket\up$ is a state
polarized along the $z$ axis.  More specifically, the state that we
can prepare has
\begin{align}
  \theta = \f{\pi}{2} + \arcsin\p{\f13},
  &&
  \alpha = \f{\pi}{3},
  &&
  \beta = -\f{2\pi}{3}\, I.
\end{align}
In order to evaluate \eqref{eq:O_const}, we therefore
expand\footnote{The expansion in \eqref{eq:polarized_proj} is acquired
  by representing an $n$-level spin by $n-1$ qubits restricted to the
  (Dicke) manifold of permutationally symmetric states.}
\begin{align}
  \bk{\mu|\theta,\alpha} = \chi_\mu e^{-i\mu\alpha},
  &&
  \chi_\mu \equiv { 2I \choose I+\mu }^{1/2}
  \cos\p{\f{\theta}{2}}^{I+\mu} \sin\p{\f{\theta}{2}}^{I-\mu},
  \label{eq:polarized_proj}
\end{align}
which implies that
\begin{align}
  \bk{\mu|\psi} = \chi_\mu \times \sqrt{2}\, \cos\p{\mu\alpha-\beta},
  &&
  r = \sum_\mu \chi_\mu^4 \times 4 \cos\p{\mu\alpha-\beta}^4.
\end{align}
Figure \ref{fig:limiting_vals} shows several values of $r$ for different $n$.

\begin{figure}
  \centering
  \includegraphics{figures/oscillations/const_vals.pdf}
  \includegraphics{figures/oscillations/var_z_vals.pdf}
  \caption{Several values of $r$ and $\var\p{s_\z/I}$ for different
    $n$.}
  \label{fig:limiting_vals}
\end{figure}

\subsection{Small-angle limit}

In the limit of a small spin-orbit coupling angle, $\phi I\ll1$, as
long as $tJ\ll\p{\phi I}^{-1}$ we can expand the Bessel function in
\eqref{eq:O_time} as $\J_0\p{\tau}^2\approx1-\tau^2/2+O\p{\tau^4}$ and
simplify the resulting expression to get
\begin{align}
  \bk{\Psi|\O\p{t}|\Psi}
  \approx 1 - \f14 \var\p{s_\z} \p{tJ\phi}^2,
  &&
  \var\p{X} \equiv \bk{\psi|X^2|\psi} - \bk{\psi|X|\psi}^2.
\end{align}
Figure \ref{fig:limiting_vals} shows several values of
$\var\p{s_\z/I}$ for different $n$.
